Waggle dance distributions quantify collective foraging in honey bee colonies: Supplementary material

### 2022-05-03 analysis/manuscript/methods.Rmd
updated final MS edits. Just last meeting to go

As the resources patches are randomly placed in the environment, the distance after which the first resource is discovered approximately follows an exponential distribution (given by $\lambda e^{- \lambda x}$). Through the feedback mechanism that the dance floor provides, the colony can, collectively, locate the most profitable resource in its environment. For randomly placed resources in a two dimensional environment the distance to the nearest point is distributed according to a Rayleigh distribution (given by $2 \lambda \pi x e^{- \pi \lambda x^2}$) (Pyke 1978). Our simulation model shows that this describes the distances at which recruits visit resources well.  Knowing the distance distributions of scout and recruit trips we then assume that the a proportion $p$ of all trips are scout trips. With this information we can specify the distributions of distances on the dance floor (see Supplementary Material for details).

@@

For the simplified the model we assumed that the number of dances depends weakly on distance and there is a sizable quality differences between resources of a decent size and that there is a sizable intensity of the high quality resource. Foragers on scouting trips are more likely to report larger distances than foragers on recruiting trips. By linearising the function that translates the profitability into the number of waggle dance run for the largest dance duration and normalising, we arrive at simplified distribution for dance durations for scouting trips:

>>>>

As the resources patches are randomly placed in the environment, the distance after which the first resource is discovered approximately follows an exponential distribution (given by $\lambda e^{- \lambda x}$). Through the feedback mechanism that the dance floor provides, the colony can, collectively, locate the most profitable resource in its environment. For randomly placed resources in a two dimensional environment the distance to the nearest point is distributed according to a Rayleigh distribution (given by $2 \lambda \pi x e^{- \pi \lambda x^2}$)) [@Pyke1978]. Our simulation model shows that this describes the distances at which recruits visit resources well.  Knowing the distance distributions of scout and recruit trips we then assume that the a proportion $p$ of all trips are scout trips. With this information we can specify the distributions of distances on the dance floor (see Supplementary Material for details).

@@

For the simplified the model we assumed that the number of dances depends weakly on distance and there is a sizable quality differences between resources of a non-negligible size and that there is a sizable intensity of the high quality resource (See Supplementary Material for detailed derivation). Foragers on scouting trips are more likely to report larger distances than foragers on recruiting trips. By linearising the function that translates the profitability into the number of waggle dance run for the largest dance duration and normalising, we arrive at simplified distribution for dance durations for scouting trips:

### 2022-05-03 analysis/manuscript/paper.Rmd
updated final MS edits. Just last meeting to go

Honeybee foraging is an extraordinary collective behaviour that is directed through the waggle dance communication system, whereby colony foraging effort is allocated through a series of feedback loops [@Frisch1993; @Seeley1995]. Here, we present a mathematical model that quantifies the extent to which honeybee colonies behave collectively in different ecological environments by inferring the proportion of dances performed by bees that have explored individually, and those that have been recruited to food sources. By applying this methodology to waggle dance data from twenty hives, we show our model closely fits real-world honeybee foraging patterns and demonstrate that colonies vary their use of waggle dance information across different landscapes. Our methodology provides a tool to identify the ecological conditions in which honeybee colonies rely on dance communication, opening the door to large-scale experimental exploration of the selection pressures that may have driven the evolution of this remarkable collective behaviour.

@@

Within honeybee colonies, a series of simple rules that determine when and how much bees perform the celebrated waggle dance (Fig. 1) mean that choices between feeding sites occur at the level of the group rather than the individual. For example, because the number of dance circuits performed by a bee on returning from a food source reflects the net energetic benefits of the trip, more of the colony's workforce will be recruited to the sweeter of two equidistant sources [@Seeley1991], or the closer of two equally sweet sources [@Hasenjager2022], without requiring any individual bee to compare options. This extraordinary system is a key example of how social insect colony behaviour can take on a form extending beyond that of the individual units [@Sumpter2006], functioning as a collective that makes decisions as a single entity.

Despite its intricacy and precision, a wealth of research now shows that foraging bees frequently do not use dance information to find forage sites [@Gruter2009]. Accordingly, in many situations, colonies that are prevented from communicating via dances achieve equal or even greater foraging success than their wild-type counterparts [@Sherman2002; @Dornhaus2004 @Dornhaus2006; @Gruter2013; @Price2019]. This apparent variation in the use of collective behaviour is intriguing because it may provide clues as to the selection pressures that were critical in the evolution of the dance communication system.

@@

Here, we present a method to determine the extent of waggle dance use in colony foraging within specific landscapes by identifying the proportion of foraging trips that are made by bees recruited through waggle dances. Our method examines the distribution of waggle run durations reported on the dance-floor and infers from its shape the relative contribution of dances for resources identified by individual search, "scouting" versus dances for resources found by following dances, “recruitment” [@Seeley1995]. Previously, the relative amount of scouting and recruitment trips have only been quantifiable by documenting foragers' individual search history, which requires tracking individual bees and is time -and labour- intensive [@Seeley1995]. Our methods provide an efficient means of quantifying collective behaviour with no requirement to manipulate hive orientations, assay foraging efficiency or document individuals' previous experience.

@@

To reveal how the patterns of waggle dance-encoded foraging distances might differ under scenarios where colonies rely on individual search versus recruitment, we simulated honeybees foraging in a landscape where resource patches were randomly placed in the environment. Foragers could locate these under two different strategies: either acting as a scout and locating resources themselves, or following a recruit strategy and locating resources by following a random dance from the dance floor [@Seeley1992] (Fig. 2a,b, details in Methods). As it is known in the simulation which individuals in the hive forage under what strategy, we can compare the distributions of foraging distances reported on the dance floor. Fig 2c shows that the shapes of the resource distance distributions for bees engaging in the two types of foraging trips are different. The distance distribution for the scout trips is akin close to that of an exponential distribution (Fig. 2c), which is the nearest neighbour distance distribution for foragers operating in a one-dimensional environment (see methods). The distribution of the distances reported for recruit trips (Fig. 2c) is a Rayleigh distribution which is the nearest-neighbour distribution in a two-dimensional environment [@Pyke1978] (see Methods).

@@

This description intrinsically captures honeybee foraging as collective decision making, whereby the foraging sites represented on the dance floor derive from a mixture of individual search and waggle dance information, modified by the profitability rule that biases recruitment towards closer or sweeter patches. The extent to which collective decision making is used is expressed in the proportion of scout dances, $p$, and the proportion of recruit dances, $1-p$. Scouting and recruiting are not fixed behavioural categories, because individual bees can engage in both over the course of their foraging lifetime, and foragers can dance on return from any successful trip irrespective of whether they were recruited to the forage site or found it individually [@Beekman2007].

In a hypothetical colony that relies only on individual search to find foraging sites, all trips are "scout" trips, but as recruitment becomes more important, the proportion of “recruit” trips will increase. By setting $p = 1$, we obtain a model based on the sole use of individual search (scouts only) and by allowing the proportion of scout trips, $p$, to take on any value between 0 and 1, we can derive the extent to which foraging occurs collectively. (Fig 2). In fitting these two models to distribution of waggle run durations decoded from real honeybee colonies we are thus able, using model selection [@Burnham2002], to infer if and, by estimating the parameter $p$, to what extent honeybee colonies use waggle dance information when foraging.

@@

For 16 out of 20 study hives, a model of collective foraging provided a better description of the data than one of individual foraging (Fig. 4a). In the other 4 sites, despite the collective model having the higher maximum likelihood, the individual model had a higher AIC value and so is more parsimonious (Supplementary Table 1). In all but one site, the collective model had a good fit (using a Kolmogorov-Smirnov statistic of $p > 0.05$, see methods) to the empirical waggle run durations (Fig. 4b), whereas the scout model was significantly different to the observed data in 8 sites (Kolmogorov-Smirnov statistic $p < 0.05$, Fig. 4b). The sites shown in Figs 4c-d are representative examples showing the model fits where the individual (Fig. 4c) and collective (Fig. 4d) models fit best. Note the closeness of the fit to the data, illustrating the overall quality of the model description.

@@

Our results show that, whilst colony-level foraging is mostly comprised of a mixture of scout and recruit foraging trips, in some circumstances, colony foraging can be better described by individual foraging alone. Thus, in some environments, the majority of foraging trips involve scouting to find new food sites rather than recruitment through dances. Note that this does not imply that these bees do not engage in dance following, because bees regularly follow dances but choose not to visit the advertised site [@Gruter2011], although it has also been shown that bees may cease dance following if it is proving unproductive [@Gruter2011; @Price2019]. Evaluating how these individual decisions influence the collective, however, has historically been a challenge as it is effectively impossible to track an individual's foraging behaviour over a landscape. As our results show that individual foraging accurately describes colony foraging in 11 different sites, and more parsimoniously than a model of collective foraging in 4 sites, these findings support the idea that individual decision making can dominate colony foraging and demonstrates further evidence for flexible waggle dance use by honey bee colonies.

@@

In the urban environment, the first principal component explained ~73% of the variance in the proportion of scouts (beta regression: $R^2$ = 0.73, $\phi$ = 10.4, p < 0.05, Fig. 5c). This component correlates positively with parks, allotments and cemeteries (8% of land cover; Table 1) and negatively with railways, however this accounts for ~1% of land cover (Table 1). The dominant land-use type by coverage, sparse residential (land coverage ~35%; Table 1), has a significant negative correlation with the first principal component for some sites removed over the jackknife sampling (Supplementary Fig. 2.), however, it is overall not a significant contributor to the variance in the loadings (Fig. 5c.). Parks, allotments and cemeteries are typically forage-rich for honeybees [@Baldock2019], a result which was not replicated here. However, note that the amount of land cover for this component is limited and the jackknifed PLS identified substantial variation in the loadings identified for the first principal components (Supplementary Fig. 3.).

@@

Here, we have presented a model to quantify the use of waggle-dance communication in collective decision-making by honeybee colonies. Whilst recruitment is well known to occur through other mechanisms besides the waggle dance, such as through olfactory cues [@Arenas2007], the exceptionally close fit of our model to waggle run durations underlines the importance of the waggle dance in honey bee foraging. The fitting of our model to waggle dance observations provides a time and labour efficient methodology to quantify collective behaviour. Further, by identifying the land-use combinations which most influence variation in the proportion of scouts, our model provides a tool to map the environment along the major axis of honeybee information use and visualise how land-use influences the use of the waggle dance.

Whilst our results show a decrease in reliance on waggle dance recruitment as resources become harder to find and foraging trips become longer [@Samuelson2021], they do not yet provide conclusive evidence for such a pattern. Our study was designed as a proof-of-principle based on an existing dataset, we did not systematically choose sites based on forage availability, and instead inferred forage availability based on land-use at each site. The dance recordings in our dataset were collected over an extended period of five months, over which time forage availability in the landscape likely changed considerably and non-uniformly across sites. We thus cannot rule out that longer term effects not captured in the data used in this study, such as resource stability [@Schurch2014], may also have contributed to our estimates of waggle dance use.

Our results within the agri-rural sites are in agreement with other studies which evaluated foraging performance with and without the dance [@Dornhaus2006; @Price2019]. These results suggest that, even though collective foraging is not always beneficial, in environments where high quality resources are present, but relatively scarce, collective foraging is being carried out by colonies, suggesting that under these conditions exchanging social information through the waggle dance confers benefits. In both agri-rural and urban landscapes we have shown that the proportion of scouting trips change with land-use (illustrated in the change in land-use shown in the maps from left to right in Fig 5). We arrived at this conclusion through the analysis of waggle dance data, using a mathematical model to fit the waggle dance durations against. This method provides a time and labour efficient methodology to quantify collective behaviour. We analysed an existing data set and thus provide a proof of concept for how this new toolkit can be used to evaluate factors influencing waggle dance recruitment.

>>>>

Honeybee foraging is an extraordinary collective behaviour that is directed through the waggle dance communication system, whereby colony foraging effort is allocated through a series of feedback loops. Here, we present a mathematical framework to explain how the intrinsic process of the waggle dance can allow a colony to exploit different landscape types. From waggle dance observations, we infer the proportion of dances performed by bees that forage individually, and those that forage collectively. By applying this methodology to waggle dance data from twenty hives, we show our model closely fits real-world honeybee foraging patterns and we demonstrate that colonies vary their use of waggle dance information across different landscapes. Our methodology provides a tool to identify the ecological conditions in which honeybee colonies rely on dance communication, opening the door to large-scale experimental exploration of the selection pressures that may have driven the evolution of this remarkable collective behaviour.

@@

No communication system is more celebrated than the honeybee waggle dance, yet there remains aspects of the behaviour which have not yet been quantified.

Within honeybee colonies, a series of simple rules determine when and how much bees perform the celebrated waggle dance (Fig. 1) mean that choices between feeding sites occur at the level of the group rather than the individual [@Seeley1994A]. For example, because the number of dance circuits performed by a bee on returning from a food source reflects the net energetic benefits of the trip, more of the colony's workforce will be recruited to the richer of two equidistant sources [@Seeley1991], or the closer of two equally rich sources [@Hasenjager2022], without requiring any individuals to go out and visit multiple resources to make a directly informed choice. This extraordinary system is a key example of how social insect colony behaviour can take on a form extending beyond that of the individual units [@Sumpter2006], functioning as a collective that makes decisions as a single entity.

Despite its intricacy and precision, a wealth of research now shows that foraging bees frequently do not use dance information to find forage sites [@Gruter2009]. Accordingly, in many situations, colonies that are prevented from communicating via dances achieve equal or even greater foraging success than their wild-type counterparts [@Sherman2002; @Dornhaus2004 @Dornhaus2006; @Gruter2013; @Price2019]. This plasticity in the use of collective behaviour is intriguing because it may provide clues as to the selection pressures that were critical in the evolution of the dance communication system.

@@

Here, we present a method to determine the extent of waggle dance use in colony foraging within specific landscapes by identifying the proportion of foraging trips that are made by bees recruited through waggle dances. Our method examines the distribution of waggle run durations reported on the dance-floor and infers from its shape the relative contribution of dances for resources identified by individual search, "scouting" versus dances for resources found by following dances, “recruitment” [@Seeley1995]. Previously, the relative amount of scouting and recruitment trips have only been quantifiable by documenting foragers' individual search history, which requires tracking individual bees and is time -and labour- intensive [@Seeley1995]. Our methods provide an efficient means of quantifying collective behaviour with no requirement to manipulate hive orientations, assay foraging efficiency or document individuals' previous experience. As such, they can identify when colonies adopt a collective foraging strategy- a critical step towards understanding when such a strategy outperforms individual-level behaviour.

@@

The patterns of waggle dance-encoded foraging distances might differ under scenarios where colonies rely on individual search versus recruitment. To show this, we simulated honeybees foraging in a landscape where resource patches were randomly placed in the environment. Foragers could locate these under two different strategies: either acting as a scout and locating resources themselves, or following a recruit strategy and locating resources by following a random dance from the dance floor [@Seeley1992] (Fig. 2a,b, details in Methods). As it is known in the simulation which individuals in the hive forage under what strategy, we can compare the distributions of foraging distances reported on the dance floor by each type of forager. Fig 2c shows that the shapes of the resource distance distributions for bees engaging in the two types of foraging trips are different. The distance distribution for the scout trips is akin close to that of an exponential distribution (Fig. 2c), which is the nearest neighbour distance distribution for foragers operating in a one-dimensional environment (see methods). The distribution of the distances reported for recruit trips (Fig. 2c) is a Rayleigh distribution which is the nearest-neighbour distribution in a two-dimensional environment [@Pyke1978] (see Methods).

@@

This description intrinsically captures honeybee foraging as collective decision making, whereby the foraging sites represented on the dance floor derive from a mixture of individual search and waggle dance information, modified by the profitability rule that biases recruitment towards closer or richer patches. The extent to which collective decision making is used is expressed in the proportion of scout dances, $p$, and the proportion of recruit dances, $1-p$. Scouting and recruiting are not fixed behavioural categories, because individual bees can engage in both over the course of their foraging lifetime, and foragers can dance on return from any successful trip irrespective of whether they were recruited to the forage site or found it individually [@Beekman2007].

We constructed two models of foraging: an individual model describing a hypothetical colony that relies only on individual search to find foraging sites, and a collective model which describes a colony using waggle dance communication to exploit resources. In the individual model, all trips are "scout" trips, but as recruitment becomes more important, the proportion of “recruit” trips will increase. By setting $p = 1$, we obtain a model based on the sole use of individual search (scouts only) and, alternatively, by allowing the proportion of scout trips, $p$, to take on any value between 0 and 1, we can model the extent to which foraging occurs collectively. (Fig 2). In fitting these two models to distribution of waggle run durations decoded from real honeybee colonies we are thus able, using model selection [@Burnham2002], to infer if and, by estimating the parameter $p$, to what extent honeybee colonies use waggle dance information when foraging.

@@

For 16 out of 20 study hives, the collective model provided a better description of the data than the individual model (Fig. 4a). In the other 4 sites, despite the collective model having the higher maximum likelihood, the individual model had a higher AIC value and so is more parsimonious (Supplementary Table 1). In all but one site, the collective model had a good fit (using a Kolmogorov-Smirnov statistic of $p > 0.05$, see methods) to the empirical waggle run durations (Fig. 4b), whereas the individual model was significantly different to the observed data in 8 sites (Kolmogorov-Smirnov statistic $p < 0.05$, Fig. 4b). The sites shown in Figs 4c-d are representative examples showing the model fits where the individual (Fig. 4c) and collective (Fig. 4d) models fit best. Note the closeness of the fit to the data, illustrating the overall quality of the model description.

@@

Our results show that, whilst colony-level foraging is mostly comprised of a mixture of scout and recruit foraging trips, in some circumstances, colony foraging can be better described by individual foraging alone. Thus, in some environments, the majority of foraging trips involve scouting to find new food sites rather than recruitment through dances. Note that this does not imply that these bees do not engage in dance following, because bees regularly follow dances but choose not to visit the advertised site [@Gruter2011], although it has also been shown that bees may cease dance following if it is proving unproductive [@Gruter2011; @Price2019]. Evaluating how these individual decisions influence the collective, however, has historically been a challenge as it is effectively impossible to track an individual's foraging behaviour over a landscape. As our results show that our model of individual foraging provides a good fit to colony foraging in 11 different sites, and provides a more parsimonious description of foraging than a model of collective foraging in 4 sites, these findings support the idea that individual decision making can dominate colony foraging and demonstrates further evidence for flexible waggle dance use by honey bee colonies.

@@

In the urban environment, the first principal component explained ~73% of the variance in the proportion of scouts (beta regression: $R^2$ = 0.73, $\phi$ = 10.4, p < 0.05, Fig. 5c). This component correlates positively with parks, allotments and cemeteries (8% of land cover; Table 1) and negatively with railways, however this accounts for ~1% of land cover (Table 1). The dominant land-use type by coverage, sparse residential (land coverage ~35%; Table 1), has a significant negative correlation with the first principal component for some sites removed over the jackknife sampling (Supplementary Fig. 2.), however, it is overall not a significant contributor to the variance in the loadings (Fig. 5c.). Parks, allotments and cemeteries are typically forage-rich for honeybees [@Baldock2019], yet our model indicates more scouting trips in these landscapes. However, note that the amount of land cover for this component is limited and the jackknifed PLS identified substantial variation in the loadings identified for the first principal components (Supplementary Fig. 3.).

@@

Wwe have presented a model to quantify the use of waggle-dance communication in collective decision-making by honeybee colonies. Whilst recruitment is well known to occur through other mechanisms besides the waggle dance, such as through olfactory cues [@Arenas2007], the exceptionally close fit of our model to waggle run durations underlines the importance of the waggle dance in honey bee foraging. The fitting of our model to waggle dance observations provides a time and labour efficient methodology to quantify collective behaviour. Our model provides a tool to map the environment along the major axis of honeybee information use and visualise how land-use influences the use of the waggle dance and thus identifies the land-use combinations which most influence variation in the proportion of scouts.

Whilst our results show a decrease in reliance on waggle dance recruitment as resources become harder to find and foraging trips become longer [@Samuelson2021], they do not yet provide conclusive evidence for such a pattern. Although the agri-rural results agree with this assessment, the urban results seem to go against this trend. As we used an existing dataset, we did not systematically choose sites based on forage availability, and instead inferred forage availability based on land-use at each site. The dance recordings in our dataset were collected over an extended period of five months, over which time forage availability in the landscape likely changed considerably and non-uniformly across sites. We thus cannot rule out that longer term effects not captured in the data used in this study, such as resource stability [@Schurch2014], may also have contributed to our estimates of waggle dance use. Our findings illustrate how our model can be used to estimate reliance on collective behaviour within different landscapes, for exploration within datasets where land-use is systematically chosen by design.

Our results within the agri-rural sites are in agreement with other studies which evaluated foraging performance with and without the dance [@Dornhaus2006; @Price2019]. These results suggest that, even though collective foraging is not always beneficial, in environments where high quality resources are present, but relatively scarce, collective foraging is being carried out by colonies, suggesting that under these conditions exchanging social information through the waggle dance confers benefits. In both agri-rural and urban landscapes we have shown that the proportion of scouting trips change with land-use (illustrated in the change in land-use shown in the maps from left to right in Fig 5). We arrived at this conclusion through the analysis of waggle dance data, using a mathematical model to fit the waggle dance durations against. This method provides a time and labour efficient methodology to quantify collective behaviour. We analysed an existing data set and thus provide a proof of concept for how this new toolkit can be used to evaluate factors influencing waggle dance recruitment. The methods described here can also provide a framework to start exploring how collective decision making can help to buffer the effects of landscape change, and in this case how flexibility in the use of the waggle dance may help to exploit resources when experiencing novel landscape mosaics such as in the highly urbanised conurbations.

### 2022-05-03 analysis/manuscript/supplementary_material.Rmd
updated final MS edits. Just last meeting to go

---
title: "Waggle dance distributions quantify collective foraging in honey bee colonies: Supplementary material"
author: "Joseph Palmer, Ash Samuelson, Elli Leadbeater and Vincent Jansen"
output:
  word_document
always_allow_html: true
bibliography: paper-refs.bib
# csl: nature.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(showtext)
showtext_auto()

@@

# add to plot
grid.arrange(arrangeGrob(allsites_plot, left = y.grob, bottom = x.grob))
```

@@

# set up proper names
# format table
df2 %>%
  arrange(site) %>%
  mutate(across(where(is.numeric), round, 3)) %>%
  rename(
    MLE = loglikelihood,
    "ΔAIC" = delta_AIC,
    D = ks_statistic,
    p_value = ks_pvalue
  ) %>%
  select(-k) %>%
  flextable() %>%
  set_caption(
    "Fit results for the fitting of the waggle dance model to each site."
  ) %>%
  set_table_properties(layout = "autofit") %>%
  fit_to_width(7.5)
```

>>>>

---
title: "Waggle dance distributions quantify collective foraging in honey bee colonies: Supplementary material"
# author: "Joseph Palmer, Ash Samuelson, Elli Leadbeater and Vincent Jansen"
output:
  word_document
always_allow_html: true
bibliography: paper-refs.bib
# csl: nature.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(showtext)
showtext_auto()

@@

# add to plot
grid.arrange(arrangeGrob(allsites_plot, left = y.grob, bottom = x.grob))

ggsave(
  "../results/figures/SI_figures/allsites_plot.png"
)

```

@@

# set up proper names
# format table
df2 %>%
  arrange(site) %>%
  mutate(across(where(is.numeric), round, 3)) %>%
  rename(
    MLE = loglikelihood,
    "ΔAIC" = delta_AIC,
    D = ks_statistic,
    p_value = ks_pvalue
  ) %>%
  select(-k) %>%
  flextable() %>%
  set_caption(
    "Fit results for the fitting of the waggle dance model to each site."
  ) %>%
  set_table_properties(layout = "autofit") %>%
  fit_to_width(7.5)

textable <- kbl(
  df2,
  caption = "Fit results for the fitting of the waggle dance model to each site.",
  booktabs = TRUE
  ) %>%
  kable_styling(latex_options = "striped")

print(textable)

print(
  xtable(
    df2, type = "latex",
    caption = "Fit results for the fitting of the waggle dance model to each site."
  ),
  scalebox='0.65',
  floating = FALSE,
  caption.placement = "top",
  file = "../results/figures/SI_figures/fittable.tex"
)

```
